### 2013-08-20 doc/dissertation/document/chapters/introduction/introduction.tex
adding my chapter 1 corrections

To achieve these goals, three complementary research and development
activities are performed. First, Monte Carlo Sythetic Acceleration
(MCSA) is applied to the neutron transport problem in a light water
reactor calculation to develop the necessary preconditioning
techniques and understanding required to solve difficult
problems. Second, using these new techniques, a series of benchmarks
for the Navier-Stokes equations are solved in different flow regimes
to research the performance of the Monte Carlo algorithm for nonlinear
problems using the new Forward-Automated Newton-MCSA algorithm
developed by this work. Third, a novel parallel algorithm based on
domain decomposed Monte Carlo methods for particle transport is
developed to fully parallelize the solution techniques developed to
solve the neutron transport and fluid flow problems on leadership
class hardware with scaling studies performed using a neutron
diffusion problem.

@@

As shown in Figure~\ref{fig:dnb_example}, many couplings are required
among individual physics components in order to accurately model this
situation with each physics generating and receiving many
responses. Those variables that are very tightly coupled, such as the
temperatures generated by the fluid dynamics and heat transfer
components, will have strong nonlinearities in their behavior and
would therefore benefit from fully consistent nonlinear solution
schemes such as that developed by this work instead of fixed-point
type iterations between physics\footnote{Fixed-point iterations
  between physics are commonly referred to as Picard iterations in the
  multiphysics community.}. To leverage such a method for analysis
such as a departure from nucleate boiling calculation, we also require
effective solutions for each of the physics involved including the
neutronics problem and the fluid problem, both of which we discuss
next.

@@

Nonlinear transport problems are a common occurrence in single and
multiple physics problems in nuclear engineering. Systems of partial
differential equations, such as the Navier-Stokes equations that
describe fluid flow, yield discrete sets of stiff equations with
nonlinearities present in the variables when discretized by
conventional methods. These sets of equations characterize momentum
and energy transport through diffusion, viscous forces and convection,
and inertial forces. Traditionally, such systems have been solved by
linearizing them in a form where the nonlinearities in the variables
are eliminated and more traditional linear methods can be used for
solutions. Often characterized as segregated methods where physics
operators are split and their action on the system approximated in
steps, such methods lack consistency and accuracy in resolving the
nonlinear component of the solution. In the last 30 years, fully
consistent nonlinear methods based on Newton's method have become more
popular and many advances have been made in the nuclear engineering
field to employ these methods.

@@

The computational resources required to solve the neutronics and fluid
dynamics problems can be tremendous. Recent work in modeling coupled
fluid flow and solid material heat and mass transfer in a reactor
subsystem, similar to the same components of the departure from
nucleate boiling example, was performed as part of analysis of the
Department of Energy's Consortium for Advanced Simulation of Light
Water Reactors (CASL) modeling and simulation hub. CASL used the
Drekar multiphysics code developed at Sandia National Laboratories
\cite{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully consistent methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Exnihilo radiation transport suite developed at Oak Ridge
National Laboratory \cite{evans_denovo:_2010}, compute trillions of
unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
meshes and $O(100,000)$ cores as well. Given the large scale and
complexity of these problems, if we aim to advance solution techniques
in neutronics and fluid flow, then we are motivated to advance the
solution of complex physics problems exploiting leadership class
levels of parallelism.

@@

The end result of these hardware changes is that the larger numbers of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
to overcome these events at the application level. With linear and
nonlinear solution schemes based on Monte Carlo techniques, such
issues are alleviated by statistical arguments. In the case of soft
failures, isolated floating point errors in Monte Carlo simulation are
absorbed within tally statistics while completely losing hardware
during a hard failure is manifested as a high variance event where
some portion of the Monte Carlo histories are lost. These stochastic
methods are a paradigm shift from current deterministic solver
techniques that will in their current form potentially suffer greatly
from the non-deterministic behavior expected from the next generation
of machines.

@@

In this work, the Forward-Automated Newton-MCSA (FANM) method is
developed as new nonlinear solution method. The key features of FANM
are: full Jacobian generation using modern Forward Automated
Differentiation (FAD) methods \cite{bartlett_automatic_2006}, and MCSA
as the inner linear solver. This method has several attractive
properties. First, the first-order approximation to the Jacobian used
in JFNK type methods is eliminated by generating the Jacobian
explicitly with the model equations through FAD. Second, the Jacobian
need not be explicitly formed by the user but is instead automated
through FAD; this eliminates the complexity of hand-coding derivatives
and has also been demonstrated to be more efficient computationally
than evaluating difference derivatives. Third, unlike GMRES, MCSA does
not build a subspace during iterations. Although the Jacobian must be
explicitly formed to use MCSA, for problems that take more than a few
GMRES iterations to converge the size of the Krylov subspace will
typically grow beyond that of the Jacobian. Finally, using MCSA to
solve the linear model provides its benefits for preconditioning,
potential parallelism, and performance.

@@

The goal of this work is to improve the iterative performance and
parallel scalability of solutions to discrete linear and nonlinear
transport problems by researching and developing a new set of domain
decomposed Monte Carlo Synthetic Acceleration methods. Improvements in
performance and parallel scalability will be demonstrated in a wide
variety of numerical experiments using both simple model neutron
transport problems and neutron transport and fluid flow problems
generated by production physics codes in both simple and complex
geometries.

>>>>

To achieve these goals, three complementary research and development
activities are performed. First, Monte Carlo Sythetic Acceleration
(MCSA) is applied to the neutron transport problem in a light water
reactor calculation to develop the necessary preconditioning
techniques and understanding required to solve difficult
problems. Second, using these new techniques, a series of benchmarks
for the Navier-Stokes equations are solved in different flow regimes
to research the performance of the Monte Carlo algorithm for nonlinear
problems using the new Forward-Automated Newton-MCSA (FANM) algorithm
developed by this work. Third, a novel parallel algorithm based on
domain decomposed Monte Carlo methods for particle transport is
developed to fully parallelize the solution techniques developed to
solve the neutron transport and fluid flow problems on leadership
class hardware with scaling studies performed in this work using a
neutron diffusion problem.

@@

As shown in Figure~\ref{fig:dnb_example}, many couplings are required
among individual physics components in order to accurately model this
situation with each physics generating and receiving many
responses. Those variables that are very tightly coupled, such as the
temperatures generated by the fluid dynamics and heat transfer
components, will have strong nonlinearities in their behavior and
would therefore benefit from fully implicit nonlinear solution schemes
such as that developed by this work instead of fixed-point type
iterations between physics\footnote{Fixed-point iterations between
  physics are commonly referred to as Picard iterations in the
  multiphysics community.}. To leverage such a method for analysis
such as a departure from nucleate boiling calculation, we also require
effective solutions for each of the physics involved including the
neutronics problem and the fluid problem, both of which we discuss
next.

@@

Nonlinear transport problems are a common occurrence in single and
multiple physics problems in nuclear engineering. Systems of partial
differential equations, such as the Navier-Stokes equations that
describe fluid flow, yield discrete sets of stiff equations with
nonlinearities present in the variables when discretized by
conventional methods. These sets of equations characterize momentum
and energy transport through diffusion, viscous forces and convection,
and inertial forces. Traditionally, such systems have been solved by
linearizing them in a form where the nonlinearities in the variables
are eliminated and more traditional linear methods can be used for
solutions. Often characterized as segregated methods where physics
operators are split and their action on the system approximated in
steps, such methods lack consistency and accuracy in resolving the
nonlinear component of the solution. In the last 30 years, fully
implicit nonlinear methods based on Newton's method have become more
popular and many advances have been made in the nuclear engineering
field to employ these methods.

@@

The computational resources required to solve the neutronics and fluid
dynamics problems can be tremendous. Recent work in modeling coupled
fluid flow and solid material heat and mass transfer in a reactor
subsystem, similar to the same components of the departure from
nucleate boiling example, was performed as part of analysis for the
Department of Energy's Consortium for Advanced Simulation of Light
Water Reactors (CASL) modeling and simulation hub. CASL used the
Drekar multiphysics code developed at Sandia National Laboratories
\cite{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully implicit methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Exnihilo radiation transport suite developed at Oak Ridge
National Laboratory \cite{evans_denovo:_2010}, compute trillions of
unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
meshes and $O(100,000)$ cores as well. Given the large scale and
complexity of these problems, if we aim to advance solution techniques
in neutronics and fluid flow for reactor analysis, then we are
motivated to advance the solution of complex physics problems
exploiting leadership class levels of parallelism.

@@

The end result of these hardware changes is that the larger numbers of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
to overcome these events at the application level. With linear and
nonlinear solution schemes based on Monte Carlo techniques, such
issues are potentially alleviated by statistical arguments. In the
case of soft failures, isolated floating point errors in Monte Carlo
simulation are absorbed within tally statistics while completely
losing hardware during a hard failure is manifested as a high variance
event where some portion of the Monte Carlo histories are lost. These
stochastic methods are a paradigm shift from current deterministic
solver techniques that will in their current form potentially suffer
greatly from the non-deterministic behavior expected from the next
generation of machines.

@@

In this work, the Forward-Automated Newton-MCSA (FANM) method is
developed as new nonlinear solution method. The key features of FANM
are: full Jacobian generation using modern Forward Automated
Differentiation (FAD) methods \cite{bartlett_automatic_2006}, and MCSA
as the inner linear solver. This method has several attractive
properties. First, the first-order approximation to the Jacobian used
in JFNK type methods is eliminated by generating the Jacobian
explicitly with the model equations through FAD. Second, the Jacobian
need not be explicitly formed by the user but is instead automated
through FAD; this eliminates the complexity of hand-coding derivatives
and has also been demonstrated to be more efficient computationally
than evaluating difference derivatives. Third, unlike GMRES, MCSA does
not build a subspace during iterations. Although the Jacobian must be
explicitly formed to use MCSA, for problems that take more than a few
GMRES iterations to converge the size of the Krylov subspace will
typically grow beyond that of the Jacobian. Finally, using MCSA to
solve the linear model provides its benefits of potential parallelism
and iterative performance.

@@

The goal of this work is to improve the iterative performance and
parallel scalability of solutions to discrete linear and nonlinear
transport problems by researching and developing a new set of domain
decomposed Monte Carlo Synthetic Acceleration methods. Improvements in
iterative performance and parallel scalability will be demonstrated in
a wide variety of numerical experiments using both simple model
neutron transport problems and neutron transport and fluid flow
problems generated by production physics codes in both simple and
complex geometries.
